---
title: "Activity Mon Analysis"
author: "TJM"
date: "February 25, 2018"
output: html_document
---

Title:  TJM file on Reprod Rsch Week 2     ####
================================================


Unzip and read in file to "Data"  
```{r setup, echo=TRUE}
setwd("C:/Users/ctmun/OneDrive/Desktop/ReproRsch")

zipF <- file.choose()    ## chose activity file in zip folder
outDir <- "C:/Users/ctmun/OneDrive/Desktop/ReproRsch"
unzip(zipF, exdir=outDir)
data <- read.csv("activity.csv", header=T)
```


Initial Explor analysis  with some output shown
head(data)
##  steps       date interval
##  1    NA 2012-10-01        0
##  2    NA 2012-10-01        5

tail(data)
##        steps       date interval
## 17563    NA 2012-11-30     2330
## 17564    NA 2012-11-30     2335

```{r, echo=TRUE}
dim(data)
##   [1] 17568     3

summary(data)
##steps                date          interval     
##Min.   :  0.00   2012-10-01:  288   Min.   :   0.0  
##1st Qu.:  0.00   2012-10-02:  288   1st Qu.: 588.8  
##Median :  0.00   2012-10-03:  288   Median :1177.5  
##Mean   : 37.38   2012-10-04:  288   Mean   :1177.5  
##3rd Qu.: 12.00   2012-10-05:  288   3rd Qu.:1766.2  
##Max.   :806.00   2012-10-06:  288   Max.   :2355.0  
##NA's   :2304     (Other)   :15840 

str(data)
## 'data.frame':	17568 obs. of  3 variables:
## $ steps   : int  NA NA NA NA NA NA NA NA NA NA ...
## $ date    : Factor w/ 61 levels "2012-10-01","2012-10-02",..: 1 1 1 1 1 1 1 1 1 1 ...
## $ interval: int  0 5 10 15 20 25 30 35 40 45 ...
```

Now some data manipulation with date and calculating total steps followed by histogram
```{r, echo=TRUE}
data$date <- as.POSIXct(data$date)

               ## Calculate and plot steps per day ##
## Calculate steps per day.  Group by date then sum steps
steps_day <- tapply(data$steps, data$date, sum, na.rm=TRUE)
steps_day <- as.data.frame(steps_day)

## plot histogram
hist(steps_day$steps_day, col = "green", xlab = "Total steps per day", main = "Histogram of Daily Total Steps")
```

Eliminate na and 0 values and calc mean and median
```{r, echo=TRUE}
ref_steps <- steps_day[!(steps_day$steps_day == 0),]
Meansteps <- mean(ref_steps)
Meansteps <- round(Meansteps, digits = 2)
## Meansteps
## [1] 10766.19
Mediansteps <- median(ref_steps)
## Mediansteps
## [1] 10765
print("The summary statistics for steps per day follow:")
print(c("Average steps per day : " , Meansteps))
print(c("Median steps per day : ", Mediansteps))
```

Calculating Daily Activity Pattern  
```{r, echo=TRUE}
library(dplyr)
Intmean <- data %>% group_by(interval) %>% summarise(meanst=mean(steps, na.rm = TRUE))
plot(Intmean$interval, Intmean$meanst, type="l", lwd=3, main = "Avg Daily steps - 5 min Intervals", xlab = "Interval", ylab = "Mean steps", col = "blue")
## Calculate max and print
a <- Intmean$meanst == max(Intmean$meanst)
b <- Intmean[a,]
b[1,2] <- round(b[1,2], digits = 2)
print(c("The maximum average step interval is:", b[1,]))
```

Impute missing values     

Tried mice package first with limited success  ## switched to manual repl with interval mean
```{r, echo=TRUE}
s <- sum(is.na(data))
## [1] 2304
print(c("The number of missing values is:", s))

## imputing with mean by interval  ###
Updated_data <- data
ExtInterval <- rep(Intmean$interval, 61)
Extmeansteps <- rep(Intmean$meanst, 61)
Extmeandata <- as.data.frame(cbind(ExtInterval, Extmeansteps))
```
Checking data and calc's
dim(Extmeandata)
[1] 17568     2
head(Extmeandata)
ExtInterval Extmeansteps
1           0    1.7169811
2           5    0.3396226
3          10    0.1320755
Extmeandata[286:290,]
ExtInterval Extmeansteps
286        2345    0.6415094
287        2350    0.2264151
288        2355    1.0754717
289           0    1.7169811
290           5    0.3396226

Now replace NA or missing values with mean for that interval

```{r, echo=TRUE}
for (i in 1:17568) {  while (is.na(Updated_data[i,1]) | Updated_data[i,1] =="") { Updated_data[i,1] <- Extmeandata[i,2]}}
## head(Updated_data)
##      steps       date interval
## 1 1.7169811 2012-10-01        0
## 2 0.3396226 2012-10-01        5
## 3 0.1320755 2012-10-01       10
## 4 0.1509434 2012-10-01       15
## 5 0.0754717 2012-10-01       20
## 6 2.0943396 2012-10-01       25
##  sum(is.na(Updated_data$steps))
##n[1] 0

## Summing data and making new histogram

newdailysteps <- tapply(Updated_data$steps, Updated_data$date, sum)
newdailysteps <- as.data.frame(newdailysteps)

### second histogram with innputed data
hist(newdailysteps$newdailysteps, col = "green", xlab = "Total steps per day", main = "Histogram of Daily Total Steps (no NA)")
```


Calculate new mean / median with Imputed data

```{r, echo=TRUE}
newmeansteps <- round(mean(newdailysteps$newdailysteps), digits=2)
## newmeansteps
## [1] 10766.19
newmediansteps <- median(newdailysteps$newdailysteps)
##  newmediansteps
##  [1] 10766.19
print("The updaated summary statistics for steps per day follow:")
print(c("New Average steps per day : " , newmeansteps))
print(c("New Median steps per day : ", newmediansteps))
```

Patterns for weekdays and weekends with panel plot 

```{r, echo=TRUE}
## calculate factor of weekday/weekend
Updated_data$date <- as.Date(Updated_data$date)
Updated_data$week <- ifelse(weekdays(Updated_data$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")

## subset dataframe into 2 separate df's. 
Weekdaydata <- Updated_data[Updated_data$week == "weekday",]
weekenddata <- Updated_data[Updated_data$week == "weekend",]
##  dim(Weekdaydata)
##  [1] 12960     4
##  dim(weekenddata)
##  [1] 4608    4

## now calculate mean steps by interval
wkdaymeans <- Weekdaydata %>% group_by(interval) %>% summarise(meanst=mean(steps, na.rm = TRUE))
wkendmeans <- weekenddata %>% group_by(interval) %>% summarise(meanst=mean(steps, na.rm = TRUE))

## panel plot with mean steps weekend and weekday  ##
par(mfrow =c(2,1), mar=c(4,4,2,1))
plot(wkdaymeans$interval, wkdaymeans$meanst, type="l", lwd=2, main="Avg Weekday steps", xlab="Interal", ylab = "Mean Steps", col="blue")
plot(wkendmeans$interval, wkendmeans$meanst, type="l", lwd=2, main="Avg Weekend steps", xlab="Interal", ylab = "Mean Steps", col="green")
```
